A Bias-Variance Decomposition of Information Transfer in Multi-Task Learning

### 2020-08-31 arxiv/transfer.tex
wip

Proposition \ref{prop_dist_transition} simplifies Theorem \ref{thm_main_informal} in the isotropic model, allowing for a more explicit statement of the bias-variance tradeoff.
Concretely, $\Psi(\beta_1, \beta)$ and $\Phi(\rho_1, \rho_2)$ corresponds to $\Delta_{\bias}$ and $\Delta_{\vari}$, respectively.
Roughly speaking, the transition threshold scales as $\frac{pd^2}{\sigma^2} - \frac{1}{\rho_1} - \frac{1}{\rho_2}$.
%We apply Proposition \ref{prop_dist_transition} to the parameter setting of Figure \ref{fig_model_shift} (the details are left to Appendix \ref{app_synthetic}).
%We can see that our result is able to predict positive or negative transfer  accurately and matches the empirical curve.
%There are several unexplained observations near the transition threshold $0$, which are caused by the concentration error $\nu$.
%We fix the target task and vary the source task, in particular the parameter $d$ which determines $\norm{\beta_1 - \beta_2}$.
%Figure \ref{fig_model_shift} shows the result.
%We observe that Proposition \ref{prop_dist_transition} explains most of the observations in Figure \ref{fig_model_shift}.
%The proof of Proposition \ref{prop_dist_transition} involves two parts.
%First, in equation \eqref{eq_te_var}, the positive variance reduction effect scales with $n_1 = \rho_1 p$, the number of source task data points.
%Second, we show that the negative effect of model-shift bias scales with $pd^2$, which is the expectation of $\norm{\beta_1 - \beta_2}^2$.
The proof of Proposition \ref{prop_dist_transition} can be found in Appendix \ref{app_proof_31}.
A key part of the analysis shows that $\hat{W}_1 / \hat{W}_2$ is roughly equal to one in the isotropic model,
thus simplifying the general condition in Theorem \ref{thm_main_informal}.

@@

In classical Rademacher or VC based theory of multi-task learning, the generalization bounds are usually presented for settings where the sample sizes are equal for all tasks \cite{B00,M06,MPR16}.
%More generally, such results are still applicable when all task data are being added simultaneously.
On the other hand, uneven sample sizes between different tasks (or even dominating tasks) have been empirically observed as a cause of negative transfer \cite{YKGLHF20}.
For such settings, we have also observed that adding more labeled data from one task does not always help.
%On the other hand, we have observed that adding more labeled data does not always improve performance in multi-task learning.
In the isotropic model, we consider what happens if we vary the source task sample size.
Our theory accurately predicts a curious phenomenon, where increasing the sample size of the source task results in negative transfer!
%Figure \ref{fig_size} provides a simulation result for such a setting.
%We observe that as $n_1 / n_2$ increases, there is a transition from positive to negative transfer.

\begin{proposition}[Source/target sample ratio]\label{prop_data_size}
	In the isotropic model, suppose that $\rho_1 > 40$ and $\rho_2 > 110$ are fixed constants, and $\Psi(\beta_1, \beta_2) > 2/(\rho_2 - 1)$.
	Then we have that
	\begin{itemize}
		\item \textbf{Positive transfer:} If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} < \frac{1}{\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 1) - \nu^{-1}}$, then w.h.p. $$\te(\hat{\beta}_2^{\MTL}) < \te(\hat{\beta}_2^{\STL}).$$
		\item \textbf{Negative transfer:} If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} > {\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 3/2) - \nu}$, then w.h.p. $$\te(\hat{\beta}_2^{\MTL}) > \te(\hat{\beta}_2^{\STL}).$$
	\end{itemize}
\end{proposition}
Proposition \ref{prop_data_size} describes the bias-variance tradeoff in terms of the sample ratio $\rho_1 / \rho_2$.
%We apply the result to the setting of Figure \ref{fig_size} (described in Appendix \ref{app_synthetic}).
%There are several unexplained observations near $y = 0$ caused by $\nu$.
The proof of Proposition \ref{prop_data_size} can be found in Appendix \ref{app_proof_32}.

@@

\noindent\textit{Example.}
	We measure covariate shift by $M = \Sigma_1^{1/2} \Sigma_2^{-1/2}$ similar to Theorem \ref{thm_main_informal}.
	We assume that the two tasks have the same ground truth model parameters, that is, $\beta_1 = \beta_2$ and $\Psi(\beta_1, \beta_2) = 0$.
	We compare two cases: (i) when $M = \id_{p\times p}$; (ii) when $M$ has $p/2$ singular values that are equal to $\lambda$ and $p/2$ singular values that are equal to $1 / \lambda$.
	Hence, $\lambda$ measures the severity of the covariate shift.
%	Figure \ref{fig_covariate} shows a simulation of this setting by varying $\lambda$.
	We observe that as source/target sample ratio increases, the performance gap between the two cases increases.

%By applying Lemma \ref{lem_cov_shift_informal}, we find that when $n_1 / n_2$ is large, having no covariate shift is the optimal choice provided that the determinant of $M^{\top}M$ is bounded.
We compare different choices of $M$ that belong to the following bounded set.
Let $\lambda_i$ be the $i$-th singular value of $M$.
Let $\mu_{\min} < \mu < \mu_{\max}$ be fixed values that do not grow with $p$.
{\begin{align*}
		\cS_{\mu}\define\bigset{M \left| \prod_{i=1}^p \lambda_i \le \mu^p, \mu_{\min} \le \lambda_i\le \mu_{\max}, \text{ for all } 1\le i\le p\right.},
\end{align*}}
%	We assume that $\beta_1$ and $\beta_2$ are generated following the isotropic model with $d = 0$.
\begin{proposition}[Covariate shift]\label{prop_covariate}
	For the two-task case, assume that $\beta_1 = \beta_2$, $\rho_1$ and $\rho_2$ are both greater than one.
	Let $g(M)$ denote the prediction loss of $\hat{\beta}_t^{\MTL}$ when $M = \Sigma_1^{1/2}\Sigma_2^{-1/2} \in\cS_{\mu}$.
	We have that
		\[ g(\mu\id) \le \bigbrace{1+ \bigo{{\rho_2}/{\rho_1}  }} \min_{M\in\cS_{\mu}} g(M). \]
\end{proposition}
Proposition \ref{prop_covariate} shows that when the sample ratio is large, having no covariate shift gives the optimal performance for multi-task learning.
The proof of Proposition \ref{prop_covariate} can be found in Appendix \ref{app_proof_33}.
%Proposition \ref{prop_covariate} implies that when $\rho_1\gg \rho_2$, having no covariate shift is the optimal choice for choosing the source task.
%This provides evidence that covariate shift is unfavorable when there are many source task datapoints,

@@

\subsection{Extension to Multiple Tasks}

>>>>

Proposition \ref{prop_dist_transition} simplifies Theorem \ref{thm_main_informal} in the isotropic model, allowing for a more explicit statement of the bias-variance tradeoff.
Concretely, $\Psi(\beta_1, \beta)$ and $\Phi(\rho_1, \rho_2)$ corresponds to $\Delta_{\bias}$ and $\Delta_{\vari}$, respectively.
%Roughly speaking, the transition threshold scales as $\frac{pd^2}{\sigma^2} - \frac{1}{\rho_1} - \frac{1}{\rho_2}$.
We apply Proposition \ref{prop_dist_transition} to the parameter setting of Figure \ref{fig_model_shift} (the details are left to Appendix \ref{app_synthetic}).
We can see that our result is able to predict positive or negative transfer  accurately and matches the empirical curve.
There are several unexplained observations near the transition threshold $0$, which are caused by the concentration error $\nu$.
%We fix the target task and vary the source task, in particular the parameter $d$ which determines $\norm{\beta_1 - \beta_2}$.
%Figure \ref{fig_model_shift} shows the result.
%We observe that Proposition \ref{prop_dist_transition} explains most of the observations in Figure \ref{fig_model_shift}.
The proof of Proposition \ref{prop_dist_transition} involves two parts.
First, in equation \eqref{eq_te_var}, the positive variance reduction effect scales with $n_1 = \rho_1 p$, the number of source task data points.
Second, we show that the negative effect of model-shift bias scales with $pd^2$, which is the expectation of $\norm{\beta_1 - \beta_2}^2$.
The proof of Proposition \ref{prop_dist_transition} can be found in Appendix \ref{app_proof_31}.
A key part of the analysis shows that $\hat{W}_1 / \hat{W}_2$ is roughly equal to one in the isotropic model,
thus simplifying the general condition in Theorem \ref{thm_main_informal}.

@@

In classical Rademacher or VC based theory of multi-task learning, the generalization bounds are usually presented for settings where the sample sizes are equal for all tasks \cite{B00,M06,MPR16}.
%More generally, such results are still applicable when all task data are being added simultaneously.
On the other hand, uneven sample sizes between different tasks (or even dominating tasks) have been empirically observed as a cause of negative transfer \cite{YKGLHF20}.
For such settings, we have also observed that adding more labeled data from one task does not always help.
%On the other hand, we have observed that adding more labeled data does not always improve performance in multi-task learning.
In the isotropic model, we consider what happens if we vary the source task sample size.
Our theory accurately predicts a curious phenomenon, where increasing the sample size of the source task results in negative transfer!
Figure \ref{fig_size} provides a simulation result for such a setting.
We observe that as $n_1 / n_2$ increases, there is a transition from positive to negative transfer.

\begin{proposition}[Source/target sample ratio]\label{prop_data_size}
	In the isotropic model, suppose that $\rho_1 > 40$ and $\rho_2 > 110$ are fixed constants, and $\Psi(\beta_1, \beta_2) > 2/(\rho_2 - 1)$.
	Then we have that
	\begin{itemize}
		\item \textbf{Positive transfer:} If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} < \frac{1}{\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 1) - \nu^{-1}}$, then w.h.p. $$\te(\hat{\beta}_2^{\MTL}) < \te(\hat{\beta}_2^{\STL}).$$
		\item \textbf{Negative transfer:} If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} > {\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 3/2) - \nu}$, then w.h.p. $$\te(\hat{\beta}_2^{\MTL}) > \te(\hat{\beta}_2^{\STL}).$$
	\end{itemize}
\end{proposition}
Proposition \ref{prop_data_size} describes the bias-variance tradeoff in terms of the sample ratio $\rho_1 / \rho_2$.
We apply the result to the setting of Figure \ref{fig_size} (described in Appendix \ref{app_synthetic}).
There are several unexplained observations near $y = 0$ caused by $\nu$.
The proof of Proposition \ref{prop_data_size} can be found in Appendix \ref{app_proof_32}.

@@

\noindent\textit{Example.}
	We measure covariate shift by $M = \Sigma_1^{1/2} \Sigma_2^{-1/2}$ similar to Theorem \ref{thm_main_informal}.
	We assume that the two tasks have the same ground truth model parameters, that is, $\beta_1 = \beta_2$ and $\Psi(\beta_1, \beta_2) = 0$.
	We compare two cases: (i) when $M = \id_{p\times p}$; (ii) when $M$ has $p/2$ singular values that are equal to $\lambda$ and $p/2$ singular values that are equal to $1 / \lambda$.
	Hence, $\lambda$ measures the severity of the covariate shift.
	Figure \ref{fig_covariate} shows a simulation of this setting by varying $\lambda$.
	We observe that as source/target sample ratio increases, the performance gap between the two cases increases.

%By applying Lemma \ref{lem_cov_shift_informal}, we find that when $n_1 / n_2$ is large, having no covariate shift is the optimal choice provided that the determinant of $M^{\top}M$ is bounded.
We compare different choices of $M$ that belong to the following bounded set.
Let $\lambda_i$ be the $i$-th singular value of $M$.
Let $\mu_{\min} < \mu < \mu_{\max}$ be fixed values that do not grow with $p$.
{\begin{align*}
		\cS_{\mu}\define\bigset{M \left| \prod_{i=1}^p \lambda_i \le \mu^p, \mu_{\min} \le \lambda_i\le \mu_{\max}, \text{ for all } 1\le i\le p\right.},
\end{align*}}
%	We assume that $\beta_1$ and $\beta_2$ are generated following the isotropic model with $d = 0$.
\begin{proposition}[Covariate shift]\label{prop_covariate}
	For the setting of the example above, assume that $\beta_1 = \beta_2$, $\rho_1$ and $\rho_2$ are both greater than one.
	Let $g(M)$ denote the prediction loss of $\hat{\beta}_t^{\MTL}$ when $M = \Sigma_1^{1/2}\Sigma_2^{-1/2} \in\cS_{\mu}$.
	We have that
		\[ g(\mu\id) \le \bigbrace{1+ \bigo{{\rho_2}/{\rho_1}  }} \min_{M\in\cS_{\mu}} g(M). \]
\end{proposition}
Proposition \ref{prop_covariate} shows that when the sample ratio is large, having no covariate shift gives the optimal performance for multi-task learning.
The proof of Proposition \ref{prop_covariate} can be found in Appendix \ref{app_proof_33}.
%Proposition \ref{prop_covariate} implies that when $\rho_1\gg \rho_2$, having no covariate shift is the optimal choice for choosing the source task.
%This provides evidence that covariate shift is unfavorable when there are many source task datapoints,

@@
